Comparing Early Evolution of Flames in X-ray Bursts in Two and Three Dimensions

### 2023-06-04 xrb3/paper.tex
finish report

In \citet{eiden:2020}, we developed the simulation methodology in
\castro\ for modeling laterally spreading flames through the accreted
layer on the surface of a neutron star.  For these models,
\castro\ evolves the compressible Euler equations with a Coriolis
force arising from rotation, constant gravity (the plane-parallel
approximation), reactive sources, and thermal diffusion (integrated
explicitly)\edit1{:
\begin{eqnarray}
\frac{\partial \rho}{\partial t} + \nabla \cdot (\rho \Ub) &=& 0 \\
\frac{\partial (\rho X_k)}{\partial t} + \nabla \cdot (\rho \Ub X_k) &=& \rho \omegadot_k \\
\frac{\partial (\rho \Ub)}{\partial t} + \nabla \cdot (\rho \Ub \Ub) + \nabla p &=& \rho {\bf g} - 2 \rho {\bf \Omega} \times \Ub \\
%% no centrifugal- \rho {\bf \Omega} \times ({\bf \Omega} \times {\bf r}) \\
\frac{\partial (\rho E)}{\partial t} + \nabla \cdot (\rho \Ub E + p \Ub) &=& \rho \Ub \cdot {\bf g} + \nabla \cdot k_\mathrm{th} \nabla T + \rho \dot{\epsilon}
%% no centrifugal - \rho (\Omegab \cdot \rb)(\Omegab \cdot \Ub) + \rho | \Omegab |^2 (\Ub \cdot r)
\end{eqnarray}
Here, $\rho$ is the mass density, $X_k$ are the mass fractions of the
nuclear species with associated creation rates $\omegadot_k$, $\Ub$
the velocity, $E$ is the specific total energy, $p$ is the pressure,
${\bf g}$ is the gravitational
acceleration, $\Omegab$ is the rotation vector, $k_\mathrm{th}$ is the thermal conductivity, and $T$ is the temperature.
}
 We use a general stellar equation of state with
radiation, ions as a perfect gas, and degenerate/relativistic
electrons described in \citet{timmes_swesty:2000} \edit1{which closes
  the system:}
\begin{eqnarray}
E &=& e + \frac{1}{2} |\Ub|^2 \\
p &=& p(\rho, e, X_k) \\
T &=& T(\rho, e, X_k)
\end{eqnarray}
Building on these
first simulations, in \citet{harpole:2021}, we explored several
different rotation rates and initial model thermal structures to
understand how the flame speed depends on the thermal structure of the
envelope.  Both of these papers used a two-dimensional axisymmetric
geometry.  The three-dimensional simulations we present here build on
these works.  We use a neutron star crust temperature of $2\times
10^8~\mathrm{K}$ and rotation rate of 1000~Hz, as followed in
\citet{harpole:2021}.  This model was chosen here since it gives a
clean well-defined flame.  The model is constructed following the same
procedure described in \citet{eiden:2020,harpole:2021}.

>>>>

In \citet{eiden:2020}, we developed the simulation methodology in
\castro\ for modeling laterally spreading flames through the accreted
layer on the surface of a neutron star.  For these models,
\castro\ evolves the compressible Euler equations with a Coriolis
force arising from rotation, constant gravity (the plane-parallel
approximation), reactive sources, and thermal diffusion (integrated
explicitly)\edit1{:
\begin{eqnarray}
\frac{\partial \rho}{\partial t} + \nabla \cdot (\rho \Ub) &=& 0 \\
\frac{\partial (\rho X_k)}{\partial t} + \nabla \cdot (\rho \Ub X_k) &=& \rho \omegadot_k \\
\frac{\partial (\rho \Ub)}{\partial t} + \nabla \cdot (\rho \Ub \Ub) + \nabla p &=& \rho {\bf g} - 2 \rho {\bf \Omega} \times \Ub \\
%% no centrifugal- \rho {\bf \Omega} \times ({\bf \Omega} \times {\bf r}) \\
\frac{\partial (\rho E)}{\partial t} + \nabla \cdot (\rho \Ub E + p \Ub) &=& \rho \Ub \cdot {\bf g} + \nabla \cdot k_\mathrm{th} \nabla T + \rho \dot{\epsilon}
%% no centrifugal - \rho (\Omegab \cdot \rb)(\Omegab \cdot \Ub) + \rho | \Omegab |^2 (\Ub \cdot r)
\end{eqnarray}
Here, $\rho$ is the mass density, $X_k$ are the mass fractions of the
nuclear species with associated creation rates $\omegadot_k$, $\Ub$
the velocity, $E$ is the specific total energy, $p$ is the pressure,
${\bf g}$ is the gravitational
acceleration, $\Omegab$ is the rotation vector, $k_\mathrm{th}$ is the thermal conductivity, and $T$ is the temperature.
}
 We use a general stellar equation of state with
radiation, ions as a perfect gas, and degenerate/relativistic
electrons described in \citet{timmes_swesty:2000} \edit1{which closes
  the system:}
\begin{eqnarray}
E &=& e + \frac{1}{2} |\Ub|^2 \\
p &=& p(\rho, e, X_k) \\
T &=& T(\rho, e, X_k)
\end{eqnarray}
Building on these
first simulations, in \citet{harpole:2021}, we explored several
different rotation rates and initial model thermal structures to
understand how the flame speed depends on the thermal structure of the
envelope.  Both of these papers used a two-dimensional axisymmetric
geometry.  The three-dimensional simulations we present here build on
these works.  We use a neutron star crust temperature of $2\times
10^8~\mathrm{K}$ and rotation rate of 1000~Hz, as followed in
\citet{harpole:2021}.  This model was chosen here since it gives a
clean well-defined flame.  The model is constructed following the same
procedure described in \citet{eiden:2020,harpole:2021}.  \edit1{We
generate a hot and a cool model, and blend them laterally to setup
an initial perturbation that is in hydrostatic equilibrium.  The
structure of these models is shown
shown in Figure~\ref{fig:initial model}.}

\begin{figure}[t]
\centering
\plotone{initial_model_paper}
\caption{\label{fig:initial model} Initial model structure for both
the hot (dashed) and cool (solid) models, showing the density
and temperature structure as well as the initial composition.}
\end{figure}
